Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-07-09 Delaneyetal2024.tex
update to formulations section

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom.
  This quantity varies with water velocity and channel width.
  In subaerial channels,  sediment transport capacity can evolve consistently because water discharge variations are accommodated by changes in flow depth, width, and water velocity.
  By contrast, subglacial channels grow at a slower rate than most water discharge variations, causing a variable relationship between channel shape, water velocity, and discharge.
  Using numerical models applied to two observed hydrographs, we evaluate the differences in sediment transport capacity variability between subaerial and subglacial channels.
  The numerical experiments confirm that subglacial sediment transport is highly non-linear with respect to water discharge.
  Yet, subglacial sediment transport capacity can vary less if water discharge variations are closer to equilibrium with subglacial channel size.
  In a dedicated numerical experiment, we show that the changing channel size results in sediment transport capacity peaking before the maximum water discharge.
  Results also indicate that high subglacial sediment transport capacities can occur across a wide range of water discharges.
  This implies that hysteresis in channel size causes a variable relationship between sediment and water discharge in a transport-limited subglacial system.
  The implications are discussed for sediment discharge from glaciers with different hydro-climatic forcings.
  We also evaluate the conditions under which subglacial sediment transport capacity could co-vary with water discharge.
  These findings can also improve interpretations of sediment discharge records in glacierized catchments.
\end{abstract}

@@

Changes in glacier dynamics and hydrology have motivated numerous recent studies of sediment transport processes in cold regions \citep[e.g.][]{li2022,vergara2022,zhang2022}.
Increases in sediment transport have been observed in Greenland \citep{bendixen2017}, the European Alps \citep{costa2017}, the Himalayas \citep{li2021}, and the Andes \citep{vergara2022}.
To accurately explain observed changes to sediment transport in glacierized catchments, the processes controlling sediment discharge and its variations with water discharge  need to be examined \citep[e.g.][]{riihimaki2005,swift2005}.

@@

In a transport-limited regime, sediment discharge is controlled by sediment transport capacity, defined as the amount of sediment the water can carry.
In both subglacial and subaerial channels, sediment transport capacity depends on the shear stress between water and the sediment it flows over \citep{shields1936,meyer1948,engelund1967}, along with the width of the channel bottom $w$ over which sediment mobilizes.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau \propto v^2.
  \end{equation}
\end{linenomath*}
% 
Following mass conservation, the mean velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

@@

Because of the above, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \citep[e.g.][]{walder1994,alley1997,tucker1997,creyts2013,beaud2018,delaney2019,hewitt2019,wickert2019}.
To date, several modeling frameworks examine subglacial sediment transport with evolving channel size \citep{creyts2013,beaud2018,delaney2019,hewitt2019}.
However, these works minimally discuss the impact of different hydrological regimes on sediment transport capacity underneath glaciers for interpreting sediment transport records.
The few explicit parameterizations of subglacial sediment transport capacity with respect to water discharge have assumed fixed channel size  \citep{alley1997}. 
These formulations demonstrate a strongly non-linear response in subglacial sediment transport capacity to water discharge.
Understanding these processes is imperative for establishing the effect of hydro-climatic conditions on subglacial sediment dynamics.
This is especially the case because sediment discharge capacity controls the mobilization and deposition of sediment, making it a fundamental aspect of subglacial sediment evacuation, especially over short timescales.

@@

This manuscript has two objectives: 1) to evaluate the variability of subglacial sediment transport capacity with evolving subglacial channel size and different hydrographs and 2) to establish whether sub-seasonal water discharge can co-vary with sediment transport capacity in subglacial systems.
We use numerical models to examine sediment transport capacity in both subglacial channels that evolve in size and subaerial channels that remain fixed.
They are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
We run a model ensemble to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial channels.
Model outputs demonstrate the specific processes that can drive variability in sediment discharge capacity in subglacial systems.
These findings indicate differences in the relationships amongst water discharge, channel geometry, and water velocity in subglacial and subaerial channels.
The manuscript then discusses the implications of variability in sediment transport capacity from glaciers and the interpretation of sediment transport records.

@@

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_i\,\frac{v_{g}^{2}}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the channel length, and $v_g=\frac{Q}{S_g}$ is the water flow speed.
% 
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the Hooke angle, \citet{hooke1990}). $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
This completes the subglacial hydraulic model which is described by the state variables $S_g$ and $\Delta h$.

@@

Note that this subaerial channel model is purely algebraic, whereas the subglacial model comprises a differential equation for the evolution of $S_g$.
Thus the channel size in the subaerial model has no history dependence on the discharge $Q_w$, whereas the subglacial one does (Equation~\ref{eq:dS_dt}).

@@

The first numerical experiment aims to compare the variability between the subglacial and subaerial model outputs to a range of channel shapes, friction factors, and water discharge variability, as the subglacial channel's cross-sectional area is allowed to evolve.
%Following the findings in Section~\ref{sect:scaling}, variability in subglacial sediment discharge capacity could approach that of subaerial systems, should water discharge variations be sufficiently small.

Across the range of parameters examined (Section\,\ref{sect:imp}), variability in all model outputs remains higher in the subglacial system compared to the subaerial one for both \alpine{} and \icesheet{} (Figure~\ref{fig:multi_run}).
In some cases, subglacial model outputs' variability is a magnitude larger than their subaerial counterparts.
Results show that variability in both subglacial and subaerial outputs decreases with smoothing time longer than approximately $1$--$5$ days (Figure~\ref{fig:multi_run}).
These smoothing timescales remove the diurnal variations in water discharge, thereby reducing variability in model outputs. 
Velocity variability in \icesheet{} is substantially smaller than \alpine{} (Figure~\ref{fig:multi_run}\,a and \, d).
This could result from the subglacial conduit being in closer equilibrium with water discharge.
Variations in \icesheet{} shear stress are comparable to \alpine due to the effects the evolution of shear stress with subglacial conduit size (Equations ~\ref{eq:dS_dt} and \ref{eq:DW};). 
The much greater water flux in \icesheet{} drives the larger width integrated shear stress variations (Figure~\ref{fig:Qw}\,c and f).

@@

Conversely, the history dependence on channel size in subglacial channels means that model outputs occur across a large range of water discharges.
For instance, in subglacial channels in \alpine{},  high water velocity values and shear stresses can occur from a low water discharge  ($\sim\,4$\,\unit{m}$^3$\,\unit{s}$^{-1}$) to the maximum water discharge at over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ (Figure~\ref{fig:Qw_vari}\,a).
In \icesheet{}, water velocities close to the seasonal mean value can occur at water discharges between roughly $150$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ and $310$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
The subglacial channel's evolving width can counteract some of these effects.
Width-integrated shear stress generally increases with water discharge, with greater rank correlation compared to water velocity or shear stress (\alpine{}, Figure~\ref{fig:Qw_vari} a--c).
Yet, even the width-integrated shear stress can vary substantially relative to water discharge.
The highest values of width-integrated shear stress occur at water discharge values ranging from roughly $11$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ to over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
The variability in width-integrated shear stress is less pronounced in the \icesheet{} scenario, where the hydrograph has less diurnal variability (Figure~\ref{fig:Qw_vari} c, f).
This results from the discharge variations being in closer equilibrium with subglacial channels compared to \alpine{} (Sections~\ref{sect:scaling} and~\ref{sect:ensemble}).

@@

@@

@@

Water discharge measurements at less than daily timescales could severely limit subglacial sediment transport models in capturing specific events.
The decreased model output variability beyond $1$--$5$ days of smoothing could result in water discharge being in equilibrium with the subglacial channel when they are in fact not (Figure~\ref{fig:multi_run}).
Over this period, water discharge variations could lead to a stronger relationship with sediment discharge capacity (Figure~\ref{fig:chan_types} d and Section ~\ref{sect:scaling}).
Particularly, if water discharge evolves slowly with respect to the subglacial channel size, then it will covary with sediment transport capacity (Figure~\ref{fig:chan_types} d).
Such a strong relationship would not represent the impact of actual shorter-term fluctuations in water discharge.
Note, in this case, that if  pipe flow is assumed, with pressurized flow, but no channel evolution, then  sediment discharge capacity would covary with water discharge (Figure~\ref{fig:chan_types} a).

@@

@@

The manuscript's second objective is to establish if water discharge varies with subglacial sediment transport capacity.
Results here suggest that, even in a transport-limited subglacial system, an incoherent relationship between water and subglacial sediment discharge could be expected in most systems.
This presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
In subglacial channels, the timing of peak sediment discharge capacity occurs before peak water discharge during an event, due to evolving channel size.
In subaerial channels, the timing of peak sediment discharge capacity and water discharge coincide.
Covariance between water discharge and sediment transport capacity could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to rarely be the case, except where basal melt dominates.

@@

\begin{acknowledgements}
  SNSF Project No. $\mathrm{PZ00P2}$\_$202024$ provided  funding for I. Delaney.
  A. Tedstone acknowledges funding from the European Research Council, award $818994$ -- CASSANDRA.
  G. King and M. J. Gevers provided useful comments on a previous version of this manuscript.
  We benefited from fruitful discussions with J. Irving.
  We thank two reviewers for constructive and critical reviews on a previous version of the manuscript.
\end{acknowledgements}
\newpage
\appendix
\section{Sensitivity of sediment transport in different drainage regimes}
\label{sect:scaling}

Here three kinds of drainage regimes are examined, namely, pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions; Figure~\ref{fig:chan_types} a), subaerial channels (Figure~\ref{fig:chan_types} c), steady-state R-channels \citep[Figure~\ref{fig:chan_types} d]{rothlisberger1972}.
Here we show the sediment transport capacity scaling with a given water discharge and hydraulic gradient for three different sediment transport capacity formulas: Meyer-Peter M\"uller  \citep[MPM; ][]{meyer1948}, Engelund and Hansen \citep[EH; ][]{engelund1967}, and Bagnold \citep{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

The following hydraulic relations are used.
The Darcy-Weisbach equation can be stated as
\begin{equation}
  \label{eq:DW}
  \Psi \propto f\frac{Q^2}{D_h S^2},
\end{equation}
with $\Psi = \Delta h / l$ the head gradient and friction factor $f$.
The shear stress and stream power are, respectively,
\begin{equation}
  \label{eq:tau-omega}
  \tau \propto f v^2 = f \left(\frac{Q}{S}\right)^2, \quad  \Omega \propto \Psi Q.
\end{equation}
%

As in Section~\ref{sect:fluv}, the subaerial channel is assumed to have a width $w$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is assumed that its width can be approximated by a relation $w \propto Q^\alpha$ (Equation~\ref{eq:wcf}) with $\alpha\in [0,1]$.
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
% 
For a steady-state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential and that $S$ adjusts in steady state with $\Psi$ and $Q$.
Note that the R-channel model presented above in Section~\ref{sect:sub_mode}, Equation~\ref{eq:dS_dt} calculates $\Psi$ from the time-evolving $S_g$ via the Darcy-Weisbach Equation~\ref{eq:dh} and thus no Shreve approximation is then needed.
% Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
% 
Pipe flow-like conditions occur when an R-channel is subjected to rapid discharge variations such that the channel cannot adjust its size.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both steady-state R-channel and pipe flow, it is assumed that $D_h \propto S^{1/2}$.

With these assumptions, the Darcy-Weisbach equation~\eqref{eq:DW} can be solved for the not-fixed quantity: $H$ for a subaerial channel, $S$ for a R-channel, and $\Psi$ for pipe-flow.
Then, using equations~\eqref{eq:tau-omega}, the shear stress and stream power can be calculated.
These results of are summarised in Table~\ref{tab:eqs1}.
Now width integrated shear stress and sediment transport can be calculated for all regimes and all transport formulas.
The results of the 12 combinations are presented in Table~\ref{tab:Qs} and also in Table~\ref{tab:Qs2} where the fractions in the exponents are approximately given by decimal numbers for ease of comparison.

Table~\ref{tab:Qs} shows the scaling of the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime.
The total transport formula EH gives a slightly stronger dependence on all variables due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others (Equation~\ref{eq:Qs_eq}).
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ and $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the transport scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{3}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

The sediment discharge capacity in the steady state R-channel scales very similarly to the subaerial channel case for all relations and virtually identically for the $\alpha=0.5$ case (Table\ref{tab:eqs1}).
However, the head gradients $\Psi$ are likely higher for comparable $Q$ in an ice-sheet marginal or alpine glacier setting than in a subaerial channel \citep{alley1997}.
Thus, sediment transport capacity is likely higher in a steady-state R-channel.

%Furthermore, R-channels will rarely operate in a steady state as variations in water discharge.
%The timescale of diurnal melt or severe rain or melt is too short for the channel to reach a steady state.
%In these cases with high water discharge variability, channels operate more like a pipe of fixed cross-section \citep[e.g.][]{gimbert2016}.
%Table~\ref{tab:Qs} shows that sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two regimes when that exponent is at most $1.7$ (Table~\ref{tab:Qs}).
%Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities.
%These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep{alley1997}.

>>>>

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom.
  This quantity varies with water velocity and channel width.
  In subaerial channels,  sediment transport capacity can evolve consistently because water discharge variations are accommodated by changes in flow depth, width, and water velocity.
  By contrast, subglacial channels grow at a slower rate than most water discharge variations, causing a variable relationship between channel shape, water velocity, and discharge.
  Using numerical models applied to two observed hydrographs and sediment transport formulations, we evaluate the differences in sediment transport capacity variability between subaerial and subglacial channels.
  The numerical experiments confirm that subglacial sediment transport is highly non-linear with respect to water discharge.
  Yet, subglacial sediment transport capacity can vary less if water discharge variations are closer to equilibrium with subglacial channel size.
  In a dedicated numerical experiment, we show that the changing channel size results in sediment transport capacity peaking before the maximum water discharge.
  The results also indicate that high subglacial sediment transport capacities can occur across a wide range of water discharges.
  This implies that hysteresis in channel size causes a variable relationship between sediment and water discharge in a transport-limited subglacial system.
  The implications are discussed for sediment discharge from glaciers with different hydro-climatic forcings.
  We also evaluate the conditions under which subglacial sediment transport capacity could co-vary with water discharge.
  These findings can also improve interpretation of sediment discharge records in glacierized catchments.
\end{abstract}

@@

Changes in glacier dynamics and hydrology have motivated numerous recent studies on sediment transport processes in cold regions \citep[e.g.][]{li2022,vergara2022,zhang2022}.
Increases in sediment transport have been observed in Greenland \citep{bendixen2017}, the European Alps \citep{costa2017}, the Himalayas \citep{li2021}, and the Andes \citep{vergara2022}.
To accurately explain observed changes in sediment transport in glacierized catchments, the processes controlling sediment discharge and its variations with water discharge  need to be examined \citep[e.g.][]{riihimaki2005,swift2005}.

@@

In a transport-limited regime, sediment discharge is controlled by sediment transport capacity, which is defined as the amount of sediment the water can carry.
In both subglacial and subaerial channels, sediment transport capacity depends on the shear stress between water and the sediment it flows over \citep{shields1936,meyer1948,engelund1967}, along with the width of the channel bottom $w$ over which sediment mobilizes.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau \propto v^2.
  \end{equation}
\end{linenomath*}
% 
Following mass conservation, the mean velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

@@

Because of the above, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \citep[e.g.][]{walder1994,alley1997,tucker1997,creyts2013,beaud2018,delaney2019,hewitt2019,wickert2019}.
To date, several modeling frameworks examine subglacial sediment transport with evolving channel size \citep{creyts2013,beaud2018,delaney2019,hewitt2019}.
However, these works minimally discuss the impact of different hydrological regimes on sediment transport capacity underneath glaciers in the context of interpreting sediment transport records.
The few explicit parameterizations of subglacial sediment transport capacity with respect to water discharge assume fixed channel size  \citep{alley1997}. 
These formulations demonstrate a strongly non-linear response in subglacial sediment transport capacity to water discharge.
Yet, the effects of evolving channels size could impact the sediment transport response. 
Understanding these processes is imperative for establishing the effect of hydro-climatic conditions on subglacial sediment dynamics.
This is especially the case because sediment discharge capacity controls the mobilization and deposition of sediment, making it a fundamental aspect of subglacial sediment evacuation, especially over short timescales.

@@

This manuscript has two objectives: 1) to evaluate the variability of subglacial sediment transport capacity with evolving subglacial channel size and different hydrographs, and 2) to establish whether sub-seasonal water discharge can co-vary with sediment transport capacity in subglacial systems.
We use numerical models to examine sediment transport capacity in both subglacial channels that evolve in size and subaerial channels that remain fixed.
They are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
We run a model ensemble to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial channels.
Model outputs demonstrate the specific processes that can drive variability in sediment discharge capacity in subglacial systems.
Lastly,  we present algebraic formulations of the sediment transport capacity response to water discharge in subaerial, steady-state R-channel, and pipe flow conditions.
These formulations extend the relationships presented in \citet{alley1997}, and illustrate the behavior of different channel types.
Findings indicate differences in the relationships amongst water discharge, channel geometry, and water velocity in subglacial and subaerial channels.
The manuscript then discusses the implications of variability in sediment transport capacity from glaciers and the interpretation of sediment transport records.

@@

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_i\,\frac{v_{g}^{2}}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the channel length, and $v_g=\frac{Q}{S_g}$ is the water velocity.
% 
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the Hooke angle, \citet{hooke1990}). $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
This completes the subglacial hydraulic model which is described by the state variables $S_g$ and $\Delta h$.

@@

Note that this subaerial channel model is purely algebraic, whereas the subglacial model comprises a differential equation for the evolution of $S_g$.
Thus, the channel size in the subaerial model has no history dependence on the discharge $Q_w$, whereas the subglacial one does (Equation~\ref{eq:dS_dt}).

@@

The first numerical experiment aims to compare the variability between the subglacial and subaerial model outputs to a range of channel shapes, friction factors, and water discharge variability.

Across the range of parameters examined (Section\,\ref{sect:imp}), variability in all model outputs remains higher in the subglacial system compared to the subaerial one for both \alpine{} and \icesheet{} (Figure~\ref{fig:multi_run}).
In some cases, subglacial model outputs' variability is a magnitude larger than their subaerial counterparts.
Variability in both subglacial and subaerial outputs decreases with smoothing time longer than approximately $1$--$5$ days (Figure~\ref{fig:multi_run}).
These smoothing timescales remove the diurnal variations in water discharge, thereby reducing variability in model outputs. 
Velocity variability in \icesheet{} is substantially smaller than \alpine{} (Figure~\ref{fig:multi_run}\,a and \, d).
This could result from the subglacial conduit being in closer equilibrium with water discharge.
Variations in \icesheet{} shear stress are comparable to \alpine due to the effects the evolution of shear stress with subglacial conduit size (Equations ~\ref{eq:dS_dt} and \ref{eq:DW};). 
The much greater water flux and thus subglacial channel size in \icesheet{} drives the larger width integrated shear stress variations (Figure~\ref{fig:Qw}\,c and f).

@@

Conversely, the history dependence on channel size in subglacial channels means that different sediment transport characteristics, such as velocity, occur across a large range of water discharges.
For instance, in subglacial channels in \alpine{},  high water velocity values and shear stresses can occur from a low water discharge  ($\sim\,4$\,\unit{m}$^3$\,\unit{s}$^{-1}$) to the maximum water discharge at over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ (Figure~\ref{fig:Qw_vari}\,a).
In \icesheet{}, water velocities close to the seasonal mean value can occur at water discharges between roughly $150$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ and $310$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
The subglacial channel's evolving width can counteract some of these effects.
Width-integrated shear stress generally increases with water discharge, with greater rank correlation compared to water velocity or shear stress (\alpine{}, Figure~\ref{fig:Qw_vari} a--c).
Yet, even the width-integrated shear stress can vary substantially relative to water discharge.
The highest values of width-integrated shear stress occur at water discharge values ranging from roughly $11$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ to over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
The variability in width-integrated shear stress is less pronounced in the \icesheet{} scenario, where the hydrograph has less diurnal variability (Figure~\ref{fig:Qw_vari} c, f).
This results from the discharge variations being in closer equilibrium with subglacial channels compared to \alpine{} (Sections~\ref{sect:scaling} and~\ref{sect:ensemble}).

@@

\FloatBarrier
\section{Sensitivity of sediment transport in different drainage regimes}
\label{sect:scaling}

Here we examine subaerial channels (Figure~\ref{fig:chan_types} c), steady-state R-channels \citep[Figure~\ref{fig:chan_types} d]{rothlisberger1972} and pipe-flow (i.e. R-channels of fixed size that do not adjust their size to discharge conditions; Figure~\ref{fig:chan_types} a).
The sediment transport capacity is scaled with a given water discharge and hydraulic gradient for three different sediment transport capacity formulas: Meyer-Peter M\"uller  \citep[MPM; ][]{meyer1948}, Engelund and Hansen \citep[EH; ][]{engelund1967}, and Bagnold \citep{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

We use the Darcy-Weisbach equation to evaluate shear stress
\begin{equation}
  \label{eq:DW}
  \Psi \propto f\frac{Q^2}{D_h S^2},
\end{equation}
with $\Psi = \Delta h / l$ the head gradient and friction factor $f$.
The shear stress and stream power are, respectively,
\begin{equation}
  \label{eq:tau-omega}
  \tau \propto f v^2 = f \left(\frac{Q}{S}\right)^2, \quad  \Omega \propto \Psi Q.
\end{equation}
%

As in Section~\ref{sect:fluv}, the subaerial channel is assumed to have a width $w$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is assumed that its width can be approximated by a relation $w \propto Q^\alpha$ (Equation~\ref{eq:wcf}) with $\alpha\in [0,1]$.
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
% 
For a steady-state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential and that $S$ adjusts in steady state with $\Psi$ and $Q$.
Note that the R-channel model used above (Section~\ref{sect:sub_mode}), Equation~\ref{eq:dS_dt} calculates $\Psi$ from the time-evolving $S_g$ via the Darcy-Weisbach Equation~\ref{eq:dh} and thus no Shreve approximation is then needed.
% Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
% 
Pipe flow-like conditions occur when an R-channel is subjected to rapid discharge variations such that the channel cannot adjust its size.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both steady-state R-channel and pipe flow, it is assumed that $D_h \propto S^{1/2}$.

With these assumptions, the Darcy-Weisbach equation~\eqref{eq:DW} can be solved for the not-fixed quantity: $H$ for a subaerial channel, $S$ for a R-channel, and $\Psi$ for pipe-flow.
Then, using equations~\eqref{eq:tau-omega}, the shear stress and stream power can be calculated.
These results of are summarised in Table~\ref{tab:eqs1}.
The results of the 12 combinations of shear stress and sediment transport capacity are presented in Table~\ref{tab:Qs} and also in Table~S1 where the fractions in the exponents are approximately given by decimal numbers for ease of comparison.

Table~\ref{tab:Qs} shows the scaling of the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas with respect to $Q$, $\Psi$ or $S$ for the three different regimes and for different $\alpha$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow regime.
The total transport formula EH gives a slightly stronger dependence on all variables due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$  for the others (Equation~\ref{eq:Qs_eq}).
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ and $w_f\,\tau_f$ used below scale only as $Q^2$ for pipe-flow, whereas the sediment transport capacity scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{3}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

The sediment discharge capacity in the steady state R-channel scales very similarly to the subaerial channel case for all relations and virtually identically for the $\alpha=\frac{1}{3}$ case (Table~\ref{tab:eqs1}).
However, the head gradients $\Psi$ are likely higher for comparable $Q$ in an ice-sheet marginal or alpine glacier setting than in a subaerial channel \citep{alley1997}.
Thus, sediment transport capacity is likely higher in a steady-state R-channel.

@@

R-channels rarely operate in a steady state with variations in water discharge, especially during severe rain or melt is too short for the channel to reach a steady state (Figures~\ref{fig:model_outs} and \ref{fig:Qw_vari}).
In these cases with high water discharge variability, channels can behave more like a pipe of fixed cross-section \citep[e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two regimes when that exponent is at most $\frac{7}{5}$.
Thus fluctuations of discharge on short timescales (on the order of a day) have the potential to cause conditions with very high sediment transport capacities.
These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep{alley1997}, when evolving subglacial channel size is not considered (Figure~\ref{fig:model_outs}).
Note that pipe flow would cause sediment discharge capacity to covary with water discharge (Figure~\ref{fig:chan_types} a).
However, this covariance does not occur with evolving channel size (Figures~\ref{fig:Qw_vari} and \ref{fig:chan_types} b).

@@

Water discharge measurements at less than daily timescales could severely limit subglacial sediment transport models in capturing specific events.
The decreased model output variability beyond $1$--$5$ days of smoothing could result in water discharge being in equilibrium with the subglacial channel when they are in fact not (Figure~\ref{fig:multi_run}).
Over this period, water discharge variations could lead to a stronger relationship with sediment discharge capacity (Figure~\ref{fig:chan_types} d and Section ~\ref{sect:scaling}).
Particularly, if water discharge evolves slowly with respect to the subglacial channel size, then it will covary with sediment transport capacity (Figure~\ref{fig:chan_types} d).
Such a strong relationship would not represent the impact of actual shorter-term fluctuations in water discharge.

@@

The channel width and size variations in subglacial channels presented here also occur in subaerial channels in response to water discharge \citep{phillips2016}, likely over decadal timescales or in response to individual extreme events \citep[e.g.]{slater2013,dean2013}.
These timescales are likely  considerably longer than the continuous changes to the subglacial channel's response time of days.
Furthermore, some observations from subaerial systems show a co-varying relationship between water discharge and sediment transport in time \citep[e.g.][]{schmidt2006,pitlick2021}.
Even so, if channel width changes occur in subaerial channels, then the values of $\alpha$ in Equation~\ref{eq:wcf} could change in time.
$\alpha$ must be less than $1$, and even this would cause the exponent on $Q$ in the  $w_g\,\tau_{g}$ relationships to remain substantially below that of pipeflow (Section ~\ref{sect:scaling}, Table\,\ref{tab:eqs1} and \ref{tab:Qs})).

@@

The manuscript's second objective is to establish if water discharge varies with subglacial sediment transport capacity.
In subglacial channels, the timing of peak sediment discharge capacity occurs before peak water discharge during an event, due to evolving channel size.
In subaerial channels, the timing of peak sediment discharge capacity and water discharge coincide.
Results here suggest that, even in a transport-limited subglacial system, an incoherent relationship between water and subglacial sediment discharge could be expected in most systems.
This presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Covariance between water discharge and sediment transport capacity could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to rarely be the case, except where basal melt dominates.

@@

\begin{acknowledgements}
  SNSF Project No. $\mathrm{PZ00P2}$\_$202024$ provided  funding for I. Delaney.
  A. Tedstone acknowledges funding from the European Research Council, award $818994$ -- CASSANDRA.
  G. King and M. J. Gevers provided useful comments on a previous version of this manuscript.
  We benefited from fruitful discussions with J. Irving.
  We thank two reviewers for constructive and critical reviews on a previous version of the manuscript.
\end{acknowledgements}
\newpage
\appendix
